---
title: "Average Number of Children by Party Identification"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load packages

```{r packages}
library(gssr)
library(survey)
library(dplyr)
library(ggplot2)
```

## Load GSS data (run once)

```{r load-data, cache=TRUE}
# Load the GSS cumulative data file
# This is cached so it doesn't reload every time you knit
data(gss_all)
```

## Parameters

```{r parameters}
# Set minimum age filter
min_age <- 45

# Set minimum year (design variables available from 1975 onwards)
min_year <- 1975

# Pooling window: how many surveys to pool together (centered)
pool_surveys <- 7
```

## Prepare and filter data

```{r prepare-data}
# Select relevant variables, filter, and create party identification categories
gss_subset <- gss_all %>%
  select(year, childs, partyid, age, vpsu, vstrat, wtssps) %>%
  filter(
    !is.na(childs), 
    !is.na(partyid), 
    !is.na(wtssps),
    !is.na(age),
    age >= min_age,
    year >= min_year,
    partyid <= 6  # Exclude "Other party" (value 7)
  ) %>%
  mutate(
    party_grouped = case_when(
      partyid %in% 0:2 ~ "Democrat",
      partyid == 3 ~ "Independent",
      partyid %in% 4:6 ~ "Republican",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(party_grouped))

# Verify data availability
cat("Total observations:", nrow(gss_subset), "\n")
cat("Year range:", min(gss_subset$year), "to", max(gss_subset$year), "\n")
cat("\nObservations by party identification:\n")
table(gss_subset$party_grouped) %>% print()
cat("\nDetailed party ID distribution:\n")
table(gss_subset$partyid) %>% print()
cat("\nObservations by year (last 10):\n")
table(gss_subset$year) %>% tail(10) %>% print()
```

## Create survey design object

```{r survey-design}
# Set option for handling strata with single PSUs
options(survey.lonely.psu = "adjust")

# Use full complex survey design with clustering and stratification
gss_design <- svydesign(
  ids = ~vpsu,
  strata = ~vstrat,
  weights = ~wtssps,
  data = gss_subset,
  nest = TRUE
)
```

## Compute estimates with pooled surveys

```{r compute-pooled-means}
if (pool_surveys > 1) {
  # Get all unique survey years in chronological order
  survey_years <- sort(unique(gss_subset$year))
  
  cat("Survey years available:", paste(survey_years, collapse = ", "), "\n")
  cat("Total number of surveys:", length(survey_years), "\n\n")
  
  # Determine pooling radius (how many surveys before/after)
  pool_radius <- floor(pool_surveys / 2)
  
  # Initialize results list
  pooled_results <- list()
  
  # For each survey year, pool surrounding surveys
  for (i in seq_along(survey_years)) {
    target_year <- survey_years[i]
    
    # Determine which survey indices to pool
    start_idx <- max(1, i - pool_radius)
    end_idx <- min(length(survey_years), i + pool_radius)
    years_to_pool <- survey_years[start_idx:end_idx]
    
    # Get data indices for these years
    pooled_data_indices <- which(gss_subset$year %in% years_to_pool)
    
    if (length(pooled_data_indices) > 0) {
      # Create subset design for pooled surveys
      pooled_design <- gss_design[pooled_data_indices, ]
      
      # Compute estimates for this pooled period
      est <- svyby(
        formula = ~childs,
        by = ~party_grouped,
        design = pooled_design,
        FUN = svymean,
        na.rm = TRUE
      )
      
      # Add year identifier
      est$year <- target_year
      est$surveys_pooled <- paste(years_to_pool, collapse = ",")
      est$n_surveys <- length(years_to_pool)
      
      pooled_results[[length(pooled_results) + 1]] <- est
    }
  }
  
  # Combine all results
  results_df <- do.call(rbind, pooled_results)
  cat(paste0("Pooling ", pool_surveys, " surveys (centered) for each estimate\n"))
  
} else {
  # No pooling - compute survey-by-survey estimates
  results <- svyby(
    formula = ~childs,
    by = ~party_grouped + year,
    design = gss_design,
    FUN = svymean,
    na.rm = TRUE
  )
  results_df <- as.data.frame(results)
  results_df$surveys_pooled <- as.character(results_df$year)
  results_df$n_surveys <- 1
  cat("No pooling - using single survey estimates\n")
}

# Show recent results
cat("\nRecent results:\n")
head(results_df %>% arrange(desc(year)), 15)
```

## Visualize trends

```{r visualize, fig.width=10, fig.height=6}
# Calculate confidence intervals
results_df <- results_df %>%
  mutate(
    ci_lower = childs - 1.96 * se,
    ci_upper = childs + 1.96 * se
  )

# Plot average number of children over time by party identification with confidence intervals
ggplot(results_df, aes(x = year, y = childs, color = party_grouped, fill = party_grouped)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(
    values = c("Democrat" = "#2166AC", "Independent" = "#762A83", "Republican" = "#B2182B"),
    name = "Party Identification"
  ) +
  scale_fill_manual(
    values = c("Democrat" = "#2166AC", "Independent" = "#762A83", "Republican" = "#B2182B"),
    name = "Party Identification"
  ) +
  labs(
    title = "Average Number of Children by Party Identification Over Time",
    subtitle = paste0(
      "General Social Survey (weighted estimates, age >= ", min_age, 
      ", years >= ", min_year,
      ifelse(pool_surveys > 1, paste0(", pooling ", pool_surveys, " surveys"), ""),
      ")\nShaded areas represent 95% confidence intervals"
    ),
    x = "Year",
    y = "Average Number of Children"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 11)
  )
```

## Summary statistics

```{r summary-stats}
# Overall averages by party identification (using most recent available data)
recent_years <- max(results_df$year, na.rm = TRUE) - 4

overall_recent <- results_df %>%
  filter(year >= recent_years) %>%
  group_by(party_grouped) %>%
  summarise(
    mean_childs = mean(childs, na.rm = TRUE),
    mean_se = mean(se, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(
  overall_recent,
  digits = 3,
  col.names = c("Party Identification", "Mean Children", "Mean Std. Error"),
  caption = paste0("Recent Average (", recent_years, "-", max(results_df$year, na.rm = TRUE), ")")
)
```

## Summary table by year

```{r summary-table}
# Create a summary table showing recent years
results_table <- results_df %>%
  filter(year >= 2000) %>%
  select(year, party_grouped, childs, se, n_surveys, surveys_pooled) %>%
  arrange(desc(year), party_grouped)

knitr::kable(
  results_table,
  digits = 3,
  col.names = c("Year", "Party ID", "Mean Children", "Std. Error", "N Surveys", "Surveys Pooled"),
  caption = paste0(
    "Average Number of Children by Party Identification (2000 onwards)",
    ifelse(pool_surveys > 1, paste0(" - pooling ", pool_surveys, " surveys"), "")
  )
)
```

## Comparison: Party ID vs Political Views

```{r comparison-note}
cat("Note: This analysis uses party identification (partyid) instead of political views (polviews).\n\n")
cat("Party identification grouping:\n")
cat("- Democrat: Strong Democrat (0), Not Very Strong Democrat (1), Independent Close to Democrat (2)\n")
cat("- Independent: Independent (Neither, No Response) (3)\n")
cat("- Republican: Independent Close to Republican (4), Not Very Strong Republican (5), Strong Republican (6)\n\n")
cat("Other Party (7) is excluded from this analysis.\n\n")
cat("This allows comparison between:\n")
cat("1. Self-reported ideology (liberal/moderate/conservative)\n")
cat("2. Self-reported party affiliation (Democrat/Independent/Republican)\n")
```

## Notes

- This analysis uses **party identification (partyid)** rather than political views (polviews)
- Party groups: Democrat (0-2), Independent (3), Republican (4-6); excludes "Other party" (7)
- Filters for respondents aged **`r min_age`** or older
- Analysis includes data from **`r min_year`** onwards (when design variables became available)
- Uses `wtssps` (the recommended post-stratification weight for 1972-2024 per GSS documentation)
- **Pooling approach**: Properly pools the raw survey data from **`r pool_surveys`** surveys (centered, accounting for irregular survey intervals) and computes a single weighted estimate on the pooled sample
- Properly accounts for the GSS complex survey design using clustering (`vpsu`) and stratification (`vstrat`)
- The `childs` variable records the number of children respondents have had