---
title: "Financial Situation Change by Political Ideology Over Time"
author: "Analysis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load packages

```{r packages}
library(gssr)
library(survey)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Load GSS data (run once)

```{r load-data, cache=TRUE}
# Load the GSS cumulative data file
data(gss_all)
```

## Parameters

```{r parameters}
# Design variables are available from 1975 onwards
min_year <- 1975

# Pooling window: how many surveys to pool together (centered)
# e.g., 5 = current survey plus two before and two after
pool_surveys <- 5
```

## Prepare and filter data

```{r prepare-data}
# Select relevant variables, filter, and create labels
gss_subset <- gss_all |>
  select(year, polviews, finalter, vpsu, vstrat, wtssps) |>
  filter(
    !is.na(polviews),
    polviews <= 7,      # valid ideology responses
    !is.na(finalter),
    finalter <= 3,      # valid finalter responses (1=better, 2=worse, 3=stayed same)
    !is.na(wtssps),
    year >= min_year
  ) |>
  mutate(
    ideology_label = case_when(
      polviews == 1 ~ "Extremely Liberal",
      polviews == 2 ~ "Liberal",
      polviews == 3 ~ "Slightly Liberal",
      polviews == 4 ~ "Moderate",
      polviews == 5 ~ "Slightly Conservative",
      polviews == 6 ~ "Conservative",
      polviews == 7 ~ "Extremely Conservative",
      TRUE ~ NA_character_
    ),
    polviews_grouped = case_when(
      polviews %in% 1:3 ~ "Liberal",
      polviews == 4 ~ "Moderate",
      polviews %in% 5:7 ~ "Conservative",
      TRUE ~ NA_character_
    ),
    finalter_label = case_when(
      finalter == 1 ~ "Better",
      finalter == 2 ~ "Worse",
      finalter == 3 ~ "Stayed Same",
      TRUE ~ NA_character_
    ),
    # Create binary indicators for better and worse
    finalter_better = ifelse(finalter == 1, 1, 0),
    finalter_worse = ifelse(finalter == 2, 1, 0),
    polviews_num = as.numeric(polviews),
    vpsu = as.numeric(vpsu),
    vstrat = as.numeric(vstrat),
    wtssps = as.numeric(wtssps),
    # Create year-specific strata to avoid collisions when pooling
    stratvar = interaction(year, vstrat, drop = TRUE)
  ) |>
  filter(!is.na(ideology_label), !is.na(finalter_label))

gss_subset$polviews_grouped <- factor(
  gss_subset$polviews_grouped,
  levels = c("Liberal", "Moderate", "Conservative")
)

# Verify data availability
cat("Total observations:", nrow(gss_subset), "\n")
cat("Year range:", min(gss_subset$year), "to", max(gss_subset$year), "\n")
cat("\nObservations by ideology group:\n")
table(gss_subset$polviews_grouped) |> print()
cat("\nObservations by financial situation:\n")
table(gss_subset$finalter_label) |> print()
```

## Create survey design object

Survey design handling follows the approach used in the regional analysis:
- Build year-specific strata (`stratvar = interaction(year, vstrat)`) to keep strata unique when pooling across years.
- Use `survey.lonely.psu = "adjust"` and `na.action = "na.pass"` to accommodate single-PSU strata created by pooling.

```{r survey-design}
options(survey.lonely.psu = "adjust")
options(na.action = "na.pass")

gss_design <- svydesign(
  ids = ~vpsu,
  strata = ~stratvar,
  weights = ~wtssps,
  data = gss_subset,
  nest = TRUE
)
```

## Compute proportions by ideology with pooled surveys

```{r compute-pooled}
# Unique survey years and pooling radius
survey_years <- sort(unique(gss_subset$year))
pool_radius <- floor(pool_surveys / 2)

cat("Survey years available:", paste(survey_years, collapse = ", "), "\n")
cat("Total number of surveys:", length(survey_years), "\n\n")

# Containers
pooled_results <- list()

ideology_groups <- levels(gss_subset$polviews_grouped)

for (i in seq_along(survey_years)) {
  target_year <- survey_years[i]

  # Determine which survey indices to pool
  start_idx <- max(1, i - pool_radius)
  end_idx <- min(length(survey_years), i + pool_radius)
  years_to_pool <- survey_years[start_idx:end_idx]

  pooled_data_indices <- which(gss_subset$year %in% years_to_pool)

  if (length(pooled_data_indices) > 0) {
    pooled_design <- gss_design[pooled_data_indices, ]

    for (ideol in ideology_groups) {
      ideology_indices <- which(gss_subset$polviews_grouped[pooled_data_indices] == ideol)

      if (length(ideology_indices) > 0) {
        ideology_design <- pooled_design[ideology_indices, ]

        # Distribution across all finalter categories
        finalter_props <- svymean(
          ~finalter_label,
          design = ideology_design,
          na.rm = TRUE
        )

        prop_values <- as.numeric(finalter_props)
        prop_names <- gsub("finalter_label", "", names(finalter_props))
        se_values <- as.numeric(SE(finalter_props))

        props_df <- data.frame(
          year = rep(target_year, length(prop_names)),
          ideology = rep(ideol, length(prop_names)),
          financial_status = prop_names,
          proportion = prop_values,
          se = se_values,
          ci_lower = prop_values - 1.96 * se_values,
          ci_upper = prop_values + 1.96 * se_values,
          n_surveys = rep(length(years_to_pool), length(prop_names)),
          stringsAsFactors = FALSE
        )

        pooled_results[[length(pooled_results) + 1]] <- props_df
      }
    }
  }
}

# Combine results
results_df <- do.call(rbind, pooled_results)

results_df$financial_status <- factor(
  results_df$financial_status,
  levels = c("Better", "Stayed Same", "Worse")
)

results_df$ideology <- factor(
  results_df$ideology,
  levels = c("Liberal", "Moderate", "Conservative")
)

cat("Pooled estimates complete\n")
head(results_df, 15)
```

## Plot: Financial situation change by ideology over time

```{r financial-plot, fig.width=12, fig.height=11}
ideology_colors <- c(
  "Liberal" = "#1f77b4",
  "Moderate" = "#7f7f7f",
  "Conservative" = "#b2182b"
)

# Create more descriptive labels for facets
results_df$outcome_label <- case_when(
  results_df$financial_status == "Better" ~ "Financial Situation Improved",
  results_df$financial_status == "Stayed Same" ~ "Financial Situation Stayed the Same",
  results_df$financial_status == "Worse" ~ "Financial Situation Worsened"
)

results_df$outcome_label <- factor(
  results_df$outcome_label,
  levels = c("Financial Situation Improved",
             "Financial Situation Stayed the Same",
             "Financial Situation Worsened")
)

ggplot(results_df, aes(x = year, y = proportion, color = ideology, fill = ideology)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.18, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.3) +
  facet_wrap(~outcome_label, ncol = 1, scales = "free_y") +
  scale_color_manual(values = ideology_colors, name = "Political Ideology") +
  scale_fill_manual(values = ideology_colors, name = "Political Ideology") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Perceived Financial Situation Change by Political Ideology",
    subtitle = paste0(
      "General Social Survey (pooling ", pool_surveys, " surveys)\n",
      "Question: 'During the last few years, has your financial situation been getting better, worse, or stayed the same?'\n",
      "Lines show survey-weighted proportions; ribbons are 95% confidence intervals"
    ),
    x = "Year",
    y = "Proportion of respondents",
    caption = "Data source: General Social Survey (GSS)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 14, face = "bold"),
    strip.text = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 11),
    plot.caption = element_text(size = 9, hjust = 0)
  )
```

## Stacked area chart alternative view

For an alternative view showing the full distribution as stacked areas:

```{r stacked-area, fig.width=12, fig.height=6}
# Stacked area chart
financial_colors <- c(
  "Better" = "#2ca02c",
  "Stayed Same" = "#ff7f0e",
  "Worse" = "#d62728"
)

ggplot(results_df, aes(x = year, y = proportion, fill = financial_status)) +
  geom_area(position = "stack", alpha = 0.8) +
  facet_wrap(~ideology, ncol = 3) +
  scale_fill_manual(values = financial_colors, name = "Financial Situation") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Full Distribution of Financial Situation Perceptions by Ideology",
    subtitle = paste0(
      "General Social Survey (pooling ", pool_surveys, " surveys)\n",
      "Stacked areas show proportion in each category within ideology group"
    ),
    x = "Year",
    y = "Proportion",
    caption = "Data source: General Social Survey (GSS)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 14, face = "bold"),
    strip.text = element_text(size = 11, face = "bold"),
    axis.title = element_text(size = 11),
    plot.caption = element_text(size = 9, hjust = 0)
  )
```

## Summary statistics by ideology (all years pooled)

```{r summary-stats}
# Compute overall proportions by ideology across all years
overall_design <- gss_design

summary_by_ideology <- list()

for (ideol in ideology_groups) {
  ideology_indices <- which(gss_subset$polviews_grouped == ideol)

  if (length(ideology_indices) > 0) {
    ideology_design <- overall_design[ideology_indices, ]

    finalter_props <- svymean(~finalter_label, design = ideology_design, na.rm = TRUE)

    prop_values <- as.numeric(finalter_props)
    prop_names <- gsub("finalter_label", "", names(finalter_props))
    se_values <- as.numeric(SE(finalter_props))

    props_df <- data.frame(
      ideology = ideol,
      financial_status = prop_names,
      proportion = prop_values,
      se = se_values,
      stringsAsFactors = FALSE
    )

    summary_by_ideology[[length(summary_by_ideology) + 1]] <- props_df
  }
}

summary_df <- do.call(rbind, summary_by_ideology)

cat("\nOverall proportions by ideology (all years, survey-weighted):\n")
print(summary_df, digits = 3)
```

## Notes

- Uses the same rolling pooled estimation and survey design handling as the regional ideology analysis (`stratvar` interaction, `nest = TRUE`, `survey.lonely.psu = "adjust"`).
- Ideology grouping follows the three-category system: Liberal (1-3), Moderate (4), Conservative (5-7).
- The `finalter` variable asks about change in financial situation over "the last few years."
- The main faceted line plot shows all three categories (improved, stayed same, worsened) with trends over time and 95% confidence intervals for each ideology group.
- The stacked area chart provides an alternative view showing the full distribution within each ideology group.
