---
title: "Average Number of Children by Political Views and Year"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Load packages

```{r packages}
library(gssr)
library(survey)
library(dplyr)
library(ggplot2)
```

## Load GSS data (run once)

```{r load-data, cache=TRUE}
# Load the GSS cumulative data file
# This is cached so it doesn't reload every time you knit
data(gss_all)
```

## Parameters

```{r parameters}
# Set minimum age filter
min_age <- 35

# Set minimum year (design variables available from 1975 onwards)
min_year <- 1975

# Set rolling average window (in years)
rolling_window <- 5
```

## Prepare and filter data

```{r prepare-data}
# Select relevant variables, filter, and create political view categories
gss_subset <- gss_all %>%
  select(year, childs, polviews, age, vpsu, vstrat, wtssps) %>%
  filter(
    !is.na(childs), 
    !is.na(polviews), 
    !is.na(wtssps),
    !is.na(age),
    age >= min_age,
    year >= min_year
  ) %>%
  mutate(
    polviews_grouped = case_when(
      polviews %in% 1:2 ~ "Liberal",
      polviews == 4 ~ "Moderate",
      polviews %in% 6:7 ~ "Conservative",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(polviews_grouped))

# Verify data availability
cat("Total observations:", nrow(gss_subset), "\n")
cat("Year range:", min(gss_subset$year), "to", max(gss_subset$year), "\n")
cat("Observations by year (last 10):\n")
table(gss_subset$year) %>% tail(10) %>% print()
```

## Create survey design object

```{r survey-design}
# Set option for handling strata with single PSUs
# "adjust" scales the variance appropriately for singleton strata
options(survey.lonely.psu = "adjust")

# Use full complex survey design with clustering and stratification
gss_design <- svydesign(
  ids = ~vpsu,
  strata = ~vstrat,
  weights = ~wtssps,
  data = gss_subset,
  nest = TRUE
)
```

## Compute weighted means by year

```{r compute-means}
# Calculate average number of children by political views and year
results <- svyby(
  formula = ~childs,
  by = ~polviews_grouped + year,
  design = gss_design,
  FUN = svymean,
  na.rm = TRUE
)

# Convert to data frame
results_df <- as.data.frame(results)
```

## Apply rolling average

```{r rolling-average}
# Function to compute trailing rolling average
compute_rolling_avg <- function(df, window) {
  df %>%
    arrange(polviews_grouped, year) %>%
    group_by(polviews_grouped) %>%
    mutate(
      childs_smooth = zoo::rollmean(childs, k = window, align = "right", fill = NA),
      se_smooth = sqrt(zoo::rollmean(se^2, k = window, align = "right", fill = NA))
    ) %>%
    ungroup()
}

# Apply rolling average if window > 1
if (rolling_window > 1) {
  if (!requireNamespace("zoo", quietly = TRUE)) {
    install.packages("zoo")
  }
  library(zoo)
  
  results_smooth <- compute_rolling_avg(results_df, rolling_window)
  cat(paste0("Applied ", rolling_window, "-year trailing moving average\n"))
} else {
  results_smooth <- results_df %>%
    mutate(childs_smooth = childs, se_smooth = se)
  cat("No smoothing applied\n")
}

# Show recent results
head(results_smooth %>% filter(!is.na(childs_smooth)) %>% arrange(desc(year)), 20)
```

## Visualize trends

```{r visualize, fig.width=10, fig.height=6}
# Plot average number of children over time by political views
plot_data <- results_smooth %>%
  filter(!is.na(childs_smooth))

ggplot(plot_data, aes(x = year, y = childs_smooth, color = polviews_grouped)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(
    values = c("Liberal" = "#2166AC", "Moderate" = "#762A83", "Conservative" = "#B2182B"),
    name = "Political Views"
  ) +
  labs(
    title = "Average Number of Children by Political Views Over Time",
    subtitle = paste0(
      "General Social Survey (weighted estimates, age >= ", min_age, 
      ", years >= ", min_year, 
      ifelse(rolling_window > 1, paste0(", ", rolling_window, "-year trailing average"), ""),
      ")"
    ),
    x = "Year",
    y = "Average Number of Children"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 11)
  )
```

## Summary statistics

```{r summary-stats}
# Overall averages by political view (using most recent available data)
recent_years <- max(results_smooth$year, na.rm = TRUE) - 4

overall_recent <- results_smooth %>%
  filter(year >= recent_years, !is.na(childs_smooth)) %>%
  group_by(polviews_grouped) %>%
  summarise(
    mean_childs = mean(childs_smooth, na.rm = TRUE),
    mean_se = mean(se_smooth, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(
  overall_recent,
  digits = 3,
  col.names = c("Political Views", "Mean Children", "Mean Std. Error"),
  caption = paste0("Recent Average (", recent_years, "-", max(results_smooth$year, na.rm = TRUE), ")")
)
```

## Summary table by year

```{r summary-table}
# Create a summary table showing recent years
results_table <- results_smooth %>%
  filter(year >= 2000, !is.na(childs_smooth)) %>%
  select(year, polviews_grouped, childs_smooth, se_smooth) %>%
  arrange(desc(year), polviews_grouped)

knitr::kable(
  results_table,
  digits = 3,
  col.names = c("Year", "Political Views", "Mean Children", "Std. Error"),
  caption = paste0(
    "Average Number of Children by Political Views (2000 onwards)",
    ifelse(rolling_window > 1, paste0(" - ", rolling_window, "-year trailing average"), "")
  )
)
```

## Notes

- This analysis filters for respondents aged **`r min_age`** or older
- Analysis includes data from **`r min_year`** onwards (when design variables became available)
- Political views are grouped as: Liberal (1-3), Moderate (4), Conservative (5-7)
- Uses `wtssps` (the recommended post-stratification weight for 1972-2024 per GSS documentation)
- **Smoothing**: A **`r rolling_window`-year trailing moving average** is applied to reduce year-to-year noise
- Properly accounts for the GSS complex survey design using clustering (`vpsu`) and stratification (`vstrat`)
- The `childs` variable records the number of children respondents have had